---
title: "5. Measuring Networks"
author: "Dai Shizuka"
date: "7/11/2018"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: false
---

Now that we have a handle on visualizing a network and the basics of relevant types of networks, we will go about the task of quantitatively describing its characteristics.
There are multiple levels at which we can measure and describe networks: 

* Node-level
* Subcomponent-level
* Networklevel

Let's start by loading the sample social network we introduced in [section 4.1](04.1_socialnetworks.html). Here, we are going to be using both the `igraph` and `asnipe` packages.

```{r, message=F, results='hide'}
library(igraph)
library(asnipe)
assoc=as.matrix(read.csv("https://dshizuka.github.io/networkanalysis/SampleData/Sample_association.csv", header=T, row.names=1))
gbi=t(assoc) 
mat=get_network(t(assoc), association_index="SRI") 
g=graph_from_adjacency_matrix(mat, "undirected", weighted=T) #create a graph object
```

```{r}
# plot the network
set.seed(10)
l=layout_with_fr(g)
plot(g, layout=l, vertex.label="", vertex.color="gold", edge.color="slateblue", edge.width=E(g)$weight*10)
```

