---
title: "8. Simulating Networks"
author: "Dai Shizuka"
date: "7/12/2018"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: false
---


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=4, fig.height=3.5, message=F) 
set.seed(2)
```

----

Here, we will simulate some social networks just to get a feel for how to create networks based on simple rules. 


```{r message=F}
library(asnipe)
library(igraph)
```

<br>

##1. Flock network with size-related variation in gregariousness

Let's imagine a scenario in which individuals vary in how often they join groups (e.g., flocks). In addition, let's imagine that individuals also vary in a trait (e.g., body size) that correlates with gregariousness--the tendency to join flocks. In this fictitious scenario, will we find correlation between the trait and centrality?

Let's start by setting up a population of 50 individuals with a trait distributed normally (i.e., mean of 20 and standard deviation of 5). We can use the `rnorm()` function to generate these random values

Note that we are going to sort the sizes so that the largest individual is listed first.
```{r}
set.seed(2)
n=50
trait=sort(rnorm(n,mean=20, sd=5), decreasing = T) 
```

Now, we'll assign gregariousness to individuals. This will be defined as an individual's probability of joining any given flock. This value, $p$, will range from 0.01 to 0.2, and will follow a uniform distribution. Again, we will sort this vector. Then, the trait and gregariousness will be correlated
```{r}
p=sort(runif(n, min=0.01, max=0.1)) #gregariousness

plot(trait, p, ylab="Gregariousness", xlab="Trait Value") #see that gregariousness and trait are correlated
```

Ok, now let's simulate some flock observations. First, we'll choose how many flocks to observe. Totally arbitrarily, let's just say we'll observe 100 flocks.
```{r}
f=100
```

Then, we'll implement a simulation that will ultimately generate an individual-by-group matrix, whose dimensions will be 50 rows x 100 columns. 
The simulation will go as follows: For each cell, the probability that the individual (rows) is found in that flock (column) depends on that individual's gregariousness value, $p$

We will first create a matrix of 0s, then fill in 1s using a loop:
```{r, fig.width=6, fig.height=6}
ibg=matrix(0,nrow=n, ncol=f)

for(i in 1:n){
  for (j in 1:f){
  ibg[i,j]=sample(c(1,0), 1, prob=c(p[i], 1-p[i]))
  }
}
```

We will then make the network using the `get_network()` function from `asnipe` to create the adjacency matrix and the `graph_from_adjacency_matrix()` function from `igraph`.
```{r, results='hide'}
adj=get_network(t(ibg), data_format="GBI", association_index="SRI")

g=graph_from_adjacency_matrix(adj, "undirected", weighted=T)
```

Let's plot the resulting network, with node size proportional to the individual's trait value.
```{r fig.height=5, fig.width=5}
plot(g,vertex.size=trait/2, vertex.label="", edge.width=E(g)$weight*5)
```

Now, let's plot the relationship between trait value and the individual's degree centrality.
```{r, fig.height=5, fig.width=5}
plot(trait, degree(g), pch=19, col="tomato")
```

As you can see, there is quite a strong correlation between trait and social position of the individual. 

```{r}
cor.test(trait, degree(g))
```

