---
title: "9. Simulating Networks"
author: "Dai Shizuka"
date: "7/12/2018"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: false
---


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=4, fig.height=3.5, message=F) 
set.seed(2)
```

----

Here, we will simulate some social networks just to get a feel for how to create networks based on simple rules. 


```{r message=F}
library(asnipe)
library(igraph)
```

##1. Flock network with size-related variation in gregariousness

Let's imagine a scenario in which individuals vary in how often they join groups (e.g., flocks). 

In addition, let's imagine that smaller individuals tend to join flocks more often than larger individuals. 

In this fictitious scenario, will we find correlation between size and centrality?

Let's start by setting up a population of 50 individuals with a normalized measure of size (i.e., size is going to be an index with a mean of 0 and standard deviation of 1).
```{r}
n=50
trait=sort(rnorm(n,mean=0))
hist(trait)
```

```{r}
n.flocks=sort(rpois(n, lambda=5))
f=length(n.flocks*2)
```

```{r}
ibg=matrix(nrow=n, ncol=f)

for(i in 1:n){
  ibg[i,sample(1:f, n.flocks[i], replace=F)]=1
}
ibg[is.na(ibg)]=0


adj=get_network(t(ibg), data_format="GBI", association_index="SRI")
g=graph_from_adjacency_matrix(adj, "undirected", weighted=T)
plot(g,vertex.size=trait+abs(min(trait))*5, vertex.label="")
```

```{r}
plot(trait, degree(g))
```